---
title: "README"
author: "Alejandra Martinez"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output:
  md_document:
    variant: markdown_github
---

Let first install package rplam.

```{r installation, results='hide', cache=FALSE, results=FALSE, warning=FALSE, comment=FALSE, message=FALSE}
library(rplam)
```

The following is a real-data example of the implemention of the robust estimator based on B-splines under a partial linear additive model.

```{r loading data}
attach(airquality)
sum(is.na(Ozone))
length(Ozone)
names(airquality)
sum(is.na(Solar.R))
sum(is.na(Wind))
sum(is.na(Temp))
sum(is.na(Month))
sum(is.na(Day))

aux <- !is.na(Solar.R)
y <- Ozone[aux]
x1 <- Temp[aux]
x2 <- Wind[aux]
x3 <- Solar.R[aux]
z <- Month[aux]

na.y <- !is.na(y)
y <- y[na.y]
x1 <- x1[na.y]
x2 <- x2[na.y]
x3 <- x3[na.y]
z <- z[na.y]

X <- cbind(x1,x2,x3)
Z <- as.matrix(z)
```

As the \code{Z} is taken as a categorical variable, we convert it into a factor variable.
```{r Z factor}
Z <- as.factor(Z)
```


The degree selected for the spline basis is 3 for the three additive functions.
```{r spline degree}
degree.spline <- 3
```

The number of internal knots selected by the BIC criteria for the classical estimator is 9.
```{r number of knots classical}
sal <- select.nknots.cl(y,Z,X,degree.spline=degree.spline)
sal$nknots
```
And so is the number of internal knots selected using the RBIC criteria used for the robust proposal.
```{r number of knots robust}
sal.rob <- select.nknots.rob(y,Z,X,degree.spline=degree.spline)
sal.rob$nknots
```



