---
title: "README"
author: "Alejandra Martinez"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output:
  md_document:
    variant: markdown_github
---

Let first install package <code>rplam</code>.

```{r installation, results='hide', cache=FALSE, results=FALSE, warning=FALSE, comment=FALSE, message=FALSE}
library(devtools)
install_github("alemermartinez/rplam")
library(rplam)
```

The following is a real-data example of the implemention of the robust estimator based on B-splines under a partial linear additive model.

```{r loading data}

data(airquality)
x <- airquality
x <- x[ complete.cases(x), ]
x <- x[, c('Ozone', 'Solar.R', 'Wind', 'Temp','Month')]
y <- as.vector(x$Ozone)
X <- as.matrix(x[, c('Solar.R', 'Wind', 'Temp')])
Z <- as.matrix(x[, 'Month'])
```

As the \code{Z} is taken as a categorical variable, we convert it into a factor variable.
```{r Z factor}
Z <- as.factor(Z)
```


The degree selected for the spline basis is 3 for the three additive functions.
```{r spline degree}
degree.spline <- 3
```

The number of internal knots selected by the BIC criteria for the classical estimator is 9.
```{r number of knots classical}
sal <- select.nknots.cl(y,Z,X,degree.spline=degree.spline)
sal$nknots
```
And so is the number of internal knots selected using the RBIC criteria used for the robust proposal.
```{r number of knots robust}
sal.rob <- select.nknots.rob(y,Z,X,degree.spline=degree.spline)
sal.rob$nknots
```



